\MakeUppercase{Layered $H$-Partitions of $k$-Planar Graphs} \thanks{This work was partly funded by NSERC and MRI.}

### 2019-05-13 k-planar.tex
kpfonts in figures

Layered $H$-partitions of small layered width in which $H$ has some additional property are a useful tool for a number of graph problems. \citet{dujmovic.joret.ea:planar} show that, if each graph $G$ in some family $\mathcal{G}$ of graphs has a layered $H$-partition for which the layered-width of the partition and the treewidth of $H$ are each upper-bounded by some constant $c_\mathcal{G}$, then each of the following quantities are upper bounded by a constant, for every $G\in\mathcal{G}$: queue number \cite{X}, track number \cite{X}, treewidth, layered treewidth \cite{X}, etc \cite{X}, etc \cite{X}, etc \cite{X}. \citet{dujmovic.esperet.ea:planar} add non-repetitive chromatic number to this list.  \citet{dujmovic.joret.ea:planar} show that two common families of graphs have such $H$-partitions: planar graphs and (more generally) bounded-genus graphs.

@@

The remainder of the paper is organized as follows: In \secref{1-planar} we prove \thmref{1-planar}.  In \secref{k-planar} we prove \thmref{k-planar}.  In \secref{consequences}, we discuss the implications of Theorems~\ref{thm:k-planar} and \ref{thm:1-planar} for other graph parameters of $k$-planar graphs.

@@

\subsubsection{Layered $H$-Partitions from PS-Trees}
\seclabel{k-to-h}

Before continuing, it is worth pointing out how an $H$-partition of $\Delta-\{r\}$ can be obtained from a PS-tree $K=K(\Delta,T,r)$. (Note: There is nothing new in this section.  This is just an alternative presentation of material from \citet{dujmovic.joret.ea:planar}.) Given $K=K(\Delta,T,r)$, we obtain a layered $H$-partition $(\mathcal{L},\mathcal{P})$ as follows.

We use the layering $\mathcal{L}=\{V_0,V_1,\ldots,V_h\}$ where $V_i=\{ v\in V(\Delta)\setminus\{r\}: \dist_{\Delta}(r,v)=i\}$ and $h=\max\{\dist_\Delta(r,v):v\in V(\Delta)\}$.  The fact that $T$ is a BFS tree ensures that $\mathcal{L}$ is, indeed, a layering of $\Delta-\{r\}$.  We use the vertex partition $\mathcal{P}=\{ S_x:=Y_x : x\in V(K)\}$.  \propref{partition} ensures that $\mathcal{P}$ is, indeed a partition of $V(\Delta)-\{r\}$.

The resulting layered $H$-partition $(\mathcal{L},\mathcal{P})$ has layered width at most 3 because, for each $x\in V(K)$, $S_x$ is made up of vertices from at most 3 vertical paths in $T$.  These are either the three vertices of $Y_{r_k}$ (if $x=r_K$) or the vertices of the open tripod $Y_x$ whose vertices are included in three vertical paths $Q_1,Q_2,Q_3$ that form the legs of the closed tripod $\overline{Y}_x$.  Therefore $|S_x\cap V_i|\le 3$.

To show that the quotient graph $H=\Delta/\mathcal{P}$ has small treewidth, we produce a tree-decomposition $\mathcal{K}=\{B_x: x\in V(K)\}$ of $H$ using the tree $K$.  For each $x\in V(K)$, the subtree $K[x]=K[\{y:x\in B_y\}]$ contains
$x$ as well as every node $y$ such that $V(Y_x)\cap V(F_y) \neq\emptyset$.
It is straightforward to verify that $K[x]$ is connected for each $x\in V(K)$. The fact that $F_x$ is a cycle in $\Delta$ and \propref{ancestor-boundary} ensures that, if $xy\in E(H)$, then one of $x$ or $y$ is a $K$-ancestor of the other. \propref{ancestor-boundary} also implies that, for any node $y\in V(K)$, $|B_y|\le 4$, since $B_y$ contains $y$ and at most 3 $K$-ancestors of $y$.

Therefore $(\mathcal{L},\mathcal{P})$ is a layered $H$-partition of $\Delta-\{r\}$ with layered-width 3 in which $H$ has treewidth 3.  Furthermore, for each $P\in\mathcal{P}$, $\Delta[P]$ is connected.  Since $\Delta$ is planar, this implies that $H$ is planar.\footnote{It is well-known and easy to see that any graph $\Delta'$ obtained by contracting an edge in a planar graph $\Delta$ is also planar.  Repeatedly applying this fact implies that $\Delta/\mathcal{P}$ is planar provided that $\Delta[P]$ is connected for every $P\in\mathcal{P}$.}

@@

  Consider the graph $A$ that contains the edges and vertices of $F_{a}$ for all $K$-ancestors $a$ of $x$. See \figref{A}.  It is helpful to think of this graph as follows:  We begin with $F_{a_0}$ where $a_0=r_K$ is the root of $K$.  We then cut $F_{a_0}$ into two pieces $F^{\bar{x}}_{a_0}$ and $F^x_{a_0}$ where $F^x_{a_0}$ is the piece that contains $F_x$. (The path that makes this cut contains two legs $Q_i$ and $Q_j$ of a tripod $Y$ and one edge $v_iv_j$ of its central triangle $\tau$.)  This process continues by partitioning $F_{a_1}:=F^x_{a_0}$ recursively.  The process terminates after some number $d$ of iterations, when $F^x_{a_{d-1}}=F_x=F_{a_d}$.

  \begin{figure}
    \begin{center}
        \includegraphics{figs/A-1}
    \end{center}
    \caption{The graph $A$.}
    \figlabel{A}
  \end{figure}

  If $B_x$ contains some $K$-ancestor $a_i$, $i\in\{0,\ldots,d-1\}$, of $x$, it is because there is an edge $vw\in E(G)$ with $v\in S_x$ and $w\in S_{a_i}$.  The edge $vw$ corresponds to a walk $W_{vw}$ in $G^+$. The walk $W_{vw}$ is contained in $N_x-F_x$, otherwise $x\not\in A_{vw}$ so $x\not\in A_v$, and $a(v)\neq x$, contradicting the fact that $v\in S_x$.

@@

  Summarizing, we have a walk $w'=v_0,v_1\ldots,v_{\ell-1},v_\ell=w^*$ in $G^+$ such that
  \begin{compactenum}[(i)]
    \item $v_0\in V(N_x-F_x)$;

@@

\noindent
\framebox{
\begin{minipage}{\textwidth}
  \note{DW}{%
    I think we need to define layered treewidth, and
    give the result of Dujmovic-Eppstein-Wood that (g,k)-planar graphs
    have O(gk)  layered treewidth. This is the best previous structural
    description of k-planar graphs. State that layered partitions are
    richer than  layered treewidth, and lead to O(1) bounds on non-rep
    chromatic number and queue-number, instead of O(log n) bounds via
    layered treewidth.

    I also think we should give explicit upper bounds on non-rep chromatic
    number for k-planar graphs, and also for queue-number (especially if
    they improve the bound in the FOCS / J.ACM paper).

@@

    We should add corollaries for d-map graphs, since they are $d^2$-planar.
    }
\end{minipage}
}
\section{$1$-Planar Graphs}
\seclabel{1-planar}

In this section, $G$ is a 1-plane graph. We will construct a   To construct a layered $H$-decomposition of $G$. First, we augment $G$ as follows:
\begin{enumerate}
  \item Add edges to $G$ to obtain a multigraph $G^+$ that is edge-maximal in the following sense:  if any two vertices $v,w\in V(G^+)$ appear on a common face\footnote{The \emph{faces} of an embedded graph $G$ are the connected components of $\R^2\setminus \bigcup_{vw\in E(G)} vw$.  We say that a vertex $v\in V(G)$ appears on a face $F$ if $v$ is contained in the closure of $F$.} $F$ of $G^+$, then there is an edge $vw\in E(G^+)$ that is contained in the boundary of $F$.

  \item Add a complete graph $K_4$ that contains $G^{+}$ in one of its inner faces, $F_0$, and then add edges to the resulting graph to obtain a triangulation $G^{++}$.
\end{enumerate}

Note that, in contrast to \secref{k-planar}, neither $G^+$ nor $G^{++}$ is necessarily a plane graph; crossings that occur in $G$ also occur in $G^+$ and $G^{++}$.

A \emph{kite} in $G^+$ is the subgraph $K=G^+[\{v,w,x,y\}]$ induced by the endpoints of a pair of crossing edges $vw,xy\in E(G^+)$.  It follows from edge-maximality that every kite is isomorphic to the complete graph $K_4$.
The edges $vw$ and $xy$ are called \emph{spars} of $K$.  The cycle $vxwy$ is called the \emph{sail} of $K$.  It follows from edge-maximality that none of the edges $vx$, $xw$, $wy$, or $yv$ are crossed by any other edges of $G^+$. Thus any edge that is a spar of a kite $K$ is not part of a sail of any kite $K'$. Observe that any spar of $K$ is incident on exactly four \emph{kite faces} of $G^+$, each of which has three edges and two vertices of $G$ on its boundary.

The 1-plane graph $G^+$ has a plane triangulation $G^-$ as a subgraph that can be obtained by removing one spar from each kite in $G^+$.  Observe that, for any spar $vw\in E(G)\setminus E(G')$ that crosses $xy\in E(G^-)$, $G^-$ contains the path $vxw$ (and $vyw$).  It follows that $\dist_{G^-}(x,y)\le 2\cdot\dist_{G^+}(x,y)$ for any pair of vertices $x,y\in V(G)$.

Our proof of \thmref{1-planar} follows quickly from the following technical lemma, which is an extension of Lemma~X in \citet{dujmovic.joret.ea:planar}.
\begin{lem}\lemlabel{induction} The setup:
  \begin{compactenum}
    \item Let $G$ and $G'$ be defined as above.
    \item Let $T$ be a BFS tree of $G'$ rooted at some vertex $r$.
    \item For each integer $i\ge 0$, let $V_i=\{v\in V(G):\dist_T(r,v)=i\}$. 
    \item Let $C$ be a cycle in $G'$ with $r$ in the exterior of $C$ and such that
    \begin{compactenum} 
      \item No edge of $C$ is crossed by any edge of $G$; and
      \item $V(C)$ can be partitioned into $P_1,\ldots,P_k$, $k\le 3$ such that for each $i\in\{1,\ldots,k\}$,
      \begin{compactenum}
        \item $C[P_i]$ is a path; and
        \item $|P_i\cap V_j| \le 15$ for all integers $j\ge 0$.
      \end{compactenum}
    \end{compactenum}
    \item Let $N$ and $N'$ be the subgraphs of $G$ and $G'$ consisting only of those edges and vertices contained in $C$ or the interior of $C$.
  \end{compactenum}
  Then $N$ has an $H$-partition $(B_x : x\in V(H))$ such that
  \begin{compactenum}
    \item $H$ is planar;
    \item for all integers $j\ge 0$, and all $x\in V(H)$, $|B_x\cap V_j|\le 15$; 
    \item for each $i\in\{1,\ldots,k\}$, there exists some $x_i\in V(H)$ such that $P_i=B_{x_i}$; and
    \item $H$ has a tree decomposition whose largest bag has size at most 4 and such that some bag contains $x_1,\ldots,x_k$.
  \end{compactenum}
\end{lem}

\begin{proof}
  The proof is by induction on the number of vertices of $N$.
  First note that $N'$ is a near-triangulation.  If $k=3$, set $R_i=P_i$ for each $i\in\{1,2,3\}$.  Otherwise, as before, split $P_1,\ldots,P_k$ to partition $V(C)$ into three sets $R_1$, $R_2$, and $R_3$ such that each $C[R_i]$ is a path and each $R_i$ contains vertices from at most one of $P_1,\ldots,P_k$.

  For each $i\in\{1,2,3\}$ and $v\in R_i$, assign the colour $i$ to $v$.
  For each $v\in V(N')$, consider the path $P_v$ that consists of the subpath of the $v$ to $r$ path in $T$ that stops at the first vertex $v'\in V(C)$. Assign $v$ to have the same colour as $v'$.

  Sperner's Lemma ensures that $G'$ contains a triangular face $\tau=v_1v_2v_3$ whose three vertices are assigned different colours. For each $i\in\{1,2,3\}$, let $Q_i=P_{v_i}$ be the path in $T$ form $v_i$ to $v_i'$.  Let $Y$ denote the subgraph of $N'$ consisting of vertices and edges $Q_1$, $Q_2$, $Q_3$, and $\tau$.  Let $Y^+$ denote the subgraph of $N$ consisting of the vertices and edges of $Y$ plus the vertices and edges of every kite formed by a crossing between an edge of $G$ and an edge of $Y$.

  We claim that, for each integer $i\ge 0$, $|V(Y^+)\cap V_i|\le 15$.  To see this, first observe that $Y$ contains at most 3 vertices of $V_i$. If a vertex $x\in V(Y^+)\setminus V(Y)$ is contained in $V_i$, then this is because $Y$ contains an edge $vw$ with $v\in V_{i'}$ and $w\in\{V_{i'+1}$ for some $i'\in\{i-1,i\}$ and $G$ contains an edge $xy$ that crosses $vw$. (Recall that $vx,xw\in E(G')$, so $\dist_{G'}(w,r)-1\le\dist_{G'}(x,r)\le\dist_{G'}(v,r)+1$.)  The graph $Y$ contains at most 6 such edges, each of which accounts for at most 2 additional vertices of $V_i$.  Therefore, in total, $|Y^+\cap V_i|\le 15$.

  Finally, let $S$ and $S^+$ denote the subgraph of $G$ containing the edges and vertices of $Y$, respectively $Y^+$, and the edges and vertices of $C$.  The graph $S^+$ has some number of bounded faces, all contained in the interior of $C$. Some of the bounded faces of $S$ are kite faces. Call the non-kite bounded faces $F_1,\ldots,F_m$ and let $C_1,\ldots,C_m$ denote their boundaries.  We claim that, for each $i\in\{1,\ldots,m\}$, if some portion of $C_i$ is contained in an edge $vw\in E(G)$ then $vw$ is not crossed by any edge of $G$.  To see this, there are three cases to consider:
  \begin{enumerate}
    \item $vw\in E(C)$. By assumption, $vw$ is not crossed by any edge of $G$.
    \item $vw\in E(Y)$. In this case, the kite containing $vw$ is in $Y^+$, so $vw$ is only incident to kite faces.
    \item $vw\in E(Y^+)\setminus E(Y)$. In this case, either $vw$ is a sail edge in which case it is not crossed by definition, or $vw$ is a spar that was added to $Y^+$ because $vw$ crosses some edge $xy\in E(Y)$.  In this latter case, $vw$ is only incident to kite faces.
  \end{enumerate}
  Therefore each $C_i$ is a cycle in $G^+$ consisting entirely of uncrossed edges. The vertices of $C_i$ can be partitioned into at most three sets $P_1'$, $P_2'$, and $P_3'$ where $P_1'\subset V(Y^+)$, $P_2'\subseteq P_a$ and $P_3'\subseteq P_b$ for some $a,b\in\{1,2,3\}$. Furthermore $C_i[P_j']$ is a path for each $j\in\{1,\ldots,3\}$. Finally, the subgraph $N_i$ of $G$ consisting of the edges and vertices of $G$ contained in $C_i$ or its interior does not contain one of the three vertices of $\tau$. Therefore, we can apply induction using the cycle $C_i$ and the partition $P_1',P_2',P_3'$ of $V(C_i)$ to obtain the desired $H$-partition and tree decomposition of $N_i$.

  The remainder of the proof finishes in the same way as the proof of Lemma~X in \cite{dujmovic.joret.ea:planar}.  $P_1,\ldots,P_k$, and $V(Y^+)\setminus V(C)$ become elements of the $H$-partition.  Elements in each of the $H$-partitions of $N_1,\ldots,N_3$ that intersect $P_1,\ldots,P_k$, or $V(Y^+)\setminus V(C)$ are discarded and all the resulting sets are combined to obtain an $H$-partition of $G$.  The desired tree decomposition of $G$ is obtained in exactly the same way as in the proof of Lemma~X in blah.

  The planarity of $H$ comes from the fact that if two edges $vw$ and $xy$ cross, then they end up in the same bag of the $H$-partition.  This means that $H$ is actually obtained by contracting connected sets of vertices in the planar graph $G'$.  It is well know that any graph obtained by contracting connected sets of vertices in a planar graph is also planar.
\end{proof}

\begin{proof}[Proof of \thmref{1-plane}]
  The same as before except that the layered width 15 from \lemref{induction} becomes 30 because \lemref{induction} uses a layering of $G'$ and distances in $G'$ can be a factor of 2 larger than in $G$.
\end{proof}

\section{Consequences}
\seclabel{consequences}

>>>>

Layered $H$-partitions of small layered width in which $H$ has some additional property are a useful tool for a number of graph problems. \citet{dujmovic.joret.ea:planar} show that, if each graph $G$ in some family $\mathcal{G}$ of graphs has a layered $H$-partition for which the layered-width of the partition and the treewidth of $H$ are each upper-bounded by some constant $c_\mathcal{G}$, then each of the following quantities are upper bounded by a constant, for every $G\in\mathcal{G}$: queue number, track number, layered treewidth, and 3-D grid drawing volume. \citet{dujmovic.esperet.ea:planar} add non-repetitive chromatic number to this list.  \citet{dujmovic.joret.ea:planar} show that two common families of graphs have such $H$-partitions: planar graphs and (more generally) bounded-genus graphs.

@@

Prior to this work, the strongest structural characterization of $k$-planar graphs was in terms of layered treewidth, which we now define.  A \emph{tree-decomposition} $\mathcal{K}$ of a graph $G$ is consists of a tree $K$ and a collection $\mathcal{K}=\{B_x:x\in V(K)\}$ of subsets of $V(G)$ indexed by nodes of $K$ such that (i)~for every $vw\in E(G)$, there exists some $x\in V(K)$ with that $v,w\in B_x$; and (ii)~for every $v\in V(G)$, $K[x] = K[\{y: x\in B_y\}]$ is connected.  A \emph{layered tree-decomposition} $(\mathcal{L},\mathcal{K})$ consists of a layering $\mathcal{L}$ and a tree-decomposition $\mathcal{K}$ of $G$. The layered-width of $(\mathcal{L},\mathcal{K})$ is $\max\{|L\cap B|: L\in \mathcal{L},\, B\in \mathcal{K}\}$.  The \emph{layered treewidth} of $G$ is the minimum layered-width of any layered tree-decomposition of $G$.

\citet{dujmovic.eppstein.ea:structure} show that $k$-planar graphs have layered treewidth $O(k+1)$.  This already implies that several of the graph parameters mentioned above, including queue-number and non-repetitive chromatic number are upper-bounded by $O(k\log n)$ for $n$-vertex $k$-planar graphs. \thmref{k-planar} is a significant strengthening of this characterization. It implies that $k$-planar graphs have layered-treewidth bounded by a function of $k$ and gives upper bounds on queue-number and non-repetitive chromatic number of $n$-vertex $k$-planar graphs that are independent of $n$.

The remainder of the paper is organized as follows: In \secref{1-planar} we prove \thmref{1-planar}.  In \secref{k-planar} we prove \thmref{k-planar}.

@@

% \subsubsection{Layered $H$-Partitions from PS-Trees}
% \seclabel{k-to-h}
% 
% Before continuing, it is worth pointing out how an $H$-partition of $\Delta-\{r\}$ can be obtained from a PS-tree $K=K(\Delta,T,r)$. (Note: There is nothing new in this section.  This is just an alternative presentation of material from \citet{dujmovic.joret.ea:planar}.) Given $K=K(\Delta,T,r)$, we obtain a layered $H$-partition $(\mathcal{L},\mathcal{P})$ as follows. 
% 
% We use the layering $\mathcal{L}=\{V_0,V_1,\ldots,V_h\}$ where $V_i=\{ v\in V(\Delta)\setminus\{r\}: \dist_{\Delta}(r,v)=i\}$ and $h=\max\{\dist_\Delta(r,v):v\in V(\Delta)\}$.  The fact that $T$ is a BFS tree ensures that $\mathcal{L}$ is, indeed, a layering of $\Delta-\{r\}$.  We use the vertex partition $\mathcal{P}=\{ S_x:=V(Y_x) : x\in V(K)\}$.  \propref{partition} ensures that $\mathcal{P}$ is, indeed a partition of $V(\Delta)-\{r\}$.
% 
% The resulting layered $H$-partition $(\mathcal{L},\mathcal{P})$ has layered width at most 3 because, for each $x\in V(K)$, $S_x$ is made up of vertices from at most 3 vertical paths in $T$.  These are either the three vertices of $Y_{r_k}$ (if $x=r_K$) or the vertices of the open tripod $Y_x$ whose vertices are included in three vertical paths $Q_1,Q_2,Q_3$ that form the legs of the closed tripod $\overline{Y}_x$.  Therefore $|S_x\cap V_i|\le 3$.
% 
% To show that the quotient graph $H=\Delta/\mathcal{P}$ has small treewidth, we produce a tree-decomposition $\mathcal{K}=\{B_x: x\in V(K)\}$ of $H$ using the tree $K$.  For each $x\in V(K)$, the subtree $K[x]=K[\{y:x\in B_y\}]$ contains
% $x$ as well as every node $y$ such that $V(Y_x)\cap V(F_y) \neq\emptyset$.
% It is straightforward to verify that $K[x]$ is connected for each $x\in V(K)$. The fact that $F_x$ is a cycle in $\Delta$ and \propref{ancestor-boundary} ensures that, if $xy\in E(H)$, then one of $x$ or $y$ is a $K$-ancestor of the other. \propref{ancestor-boundary} also implies that, for any node $y\in V(K)$, $|B_y|\le 4$, since $B_y$ contains $y$ and at most 3 $K$-ancestors of $y$.
% 
% Therefore $(\mathcal{L},\mathcal{P})$ is a layered $H$-partition of $\Delta-\{r\}$ with layered-width 3 in which $H$ has treewidth 3.  Furthermore, for each $P\in\mathcal{P}$, $\Delta[P]$ is connected.  Since $\Delta$ is planar, this implies that $H$ is planar.\footnote{It is well-known and easy to see that any graph $\Delta'$ obtained by contracting an edge in a planar graph $\Delta$ is also planar.  Repeatedly applying this fact implies that $\Delta/\mathcal{P}$ is planar provided that $\Delta[P]$ is connected for every $P\in\mathcal{P}$.}

@@

  Consider the graph $A$ that contains the edges and vertices of $F_{a}$ for all $K$-ancestors $a$ of $x$. See \figref{A}(a).  It is helpful to think of this graph as follows:  We begin with $F_{a_0}$ where $a_0=r_K$ is the root of $K$.  We then cut $F_{a_0}$ into two pieces $F^{\bar{x}}_{a_0}$ and $F^x_{a_0}$ where $F^x_{a_0}$ is the piece that contains $F_x$. (The path that makes this cut contains two legs $Q_i$ and $Q_j$ of a tripod $Y$ and one edge $v_iv_j$ of its central triangle $\tau$.)  This process continues by partitioning $F_{a_1}:=F^x_{a_0}$ recursively.  The process terminates after some number $d$ of iterations, when $F^x_{a_{d-1}}=F_x=F_{a_d}$.

  \begin{figure}

    \begin{center}
      \begin{tabular}{cc}
        \includegraphics{figs/A-1} &
        \includegraphics{figs/A-3} \\ (a) & (b)
      \end{tabular}
    \end{center}
    \caption{(a)~the graph $A$ and (b)~the walks $W_{vw}$ and $W_{ww'}$ that contain the walk $w=v_0,v_1\ldots,v_{\ell-1},v_\ell=w^*$.}
    \figlabel{A}
  \end{figure}

  Refer to \figref{A}(b)
  If $B_x$ contains some $K$-ancestor $a_i$, $i\in\{0,\ldots,d-1\}$, of $x$, it is because there is an edge $vw\in E(G)$ with $v\in S_x$ and $w\in S_{a_i}$.  The edge $vw$ corresponds to a walk $W_{vw}$ in $G^+$. The walk $W_{vw}$ is contained in $N_x-F_x$, otherwise $x\not\in A_{vw}$ so $x\not\in A_v$, and $a(v)\neq x$, contradicting the fact that $v\in S_x$.

@@

  Summarizing, we have a walk $w=v_0,v_1\ldots,v_{\ell-1},v_\ell=w^*$ in $G^+$ such that
  \begin{compactenum}[(i)]
    \item $v_0\in V(N_x-F_x)$;

@@

\noindent
\framebox{
\begin{minipage}{\textwidth}
  \note{DW}{%
    % I think we need to define layered treewidth, and
    % give the result of Dujmovic-Eppstein-Wood that (g,k)-planar graphs
    % have O(gk)  layered treewidth. This is the best previous structural
    % description of k-planar graphs. State that layered partitions are
    % richer than  layered treewidth, and lead to O(1) bounds on non-rep
    % chromatic number and queue-number, instead of O(log n) bounds via
    % layered treewidth.
    I also think we should give explicit upper bounds on non-rep chromatic
    number for k-planar graphs, and also for queue-number (especially if
    they improve the bound in the FOCS / J.ACM paper).

@@

    We should add corollaries for d-map graphs, since they are $d^2$-planar.
    }

    \note{PM}{
    A couple of other things to look into: 
    \begin{compactenum}
      \item $p$-centered colourings
      \item low treewidth colourings
    \end{compactenum}
    }
\end{minipage}
}
\section{$1$-Planar Graphs}
\seclabel{1-planar}

Let $G$ be an edge-maximal 1-plane multigraph.  Here, edge-maximal should be taken to mean that, if any two vertices $v$ and $w$ appear on a common face\footnote{The \emph{faces} of an embedded graph $G$ are the connected components of $\R^2\setminus \bigcup_{vw\in E(G)} vw$.  We say that a vertex $v\in V(G)$ appears on a face $F$ if $v$ is contained in the closure of $F$.} $F$, then there is an edge $vw\in E(G)$ that is contained in the boundary of $F$.

A \emph{kite} in $G$ is the subgraph $K=G[\{v,w,x,y\}]$ induced by the endpoints of a pair of crossing edges $vw,xy\in E(G)$.  It follows from edge-maximality that every kite is isomorphic to the complete graph $K_4$.
The edges $vw$ and $xy$ are called \emph{spars} of $K$.  The cycle $vxwy$ is called the \emph{sail} of $K$.  It follows from edge-maximality that none of the edges $vx$, $xw$, $wy$, or $yv$ are crossed by any other edges of $G$. Thus any edge that is a spar of a kite $K$ is not part of a sail of any kite $K'$. Observe that any spar of $K$ is incident on exactly four \emph{kite faces} of $G$, each of which has three edges and two vertices of $G$ on its boundary.

The 1-plane graph $G$ has a plane triangulation $G'$ as a subgraph that can be obtained by removing one spar from each kite in $G$.  Observe that, for any spar $xy\in E(G-G')$ that crosses $vw\in E(G')$, $G'$ contains the path $vxw$ (and $vyw$).  It follows that 
\begin{equation}
  \dist_{G'}(v,w)\le 2 \enspace . \eqlabel{spar-distance}
\end{equation}

Our proof of \thmref{1-planar} follows quickly from the following technical lemma, which is an extension of the equivalent result for plane graphs \cite{dujmovic.joret.ea:planar}.
\begin{lem}\lemlabel{induction} The setup:
  \begin{compactenum}
    \item Let $G$ and $G'$ be defined as above.
    \item Let $T$ be a BFS tree of $G'$ rooted at some vertex $r$.
    \item For each integer $j\ge 0$, let $V_j=\{v\in V(G):\dist_T(r,v)=j\}$. 
    \item Let $F$ be a cycle in $G'$ with $r$ in the exterior of $F$ and such that
    \begin{compactenum} 
      \item No edge of $F$ is crossed by any edge of $G$; and
      \item $V(F)$ can be partitioned into $P_1,\ldots,P_k$, $k\le 3$ such that for each $i\in\{1,\ldots,k\}$,
      \begin{compactenum}
        \item $F[P_i]$ is a path; and
        \item $|V(P_i)\cap V_j| \le 15$ for all integers $j\ge 0$.
      \end{compactenum}
    \end{compactenum}
    \item Let $N^+$ and $N^-$ be the subgraphs of $G$ and $G'$ consisting only of those edges and vertices contained in $F$ or the interior of $F$.
  \end{compactenum}
  Then $N$ has an $H$-partition $\{S_x : x\in V(H)\}$ such that
  \begin{compactenum}
    \item $H$ is planar;
    \item for all integers $j\ge 0$, and all $x\in V(H)$, $|S_x\cap V_j|\le 15$; 
    \item for each $i\in\{1,\ldots,k\}$, there exists some $x_i\in V(H)$ such that $P_i=S_{x_i}$; and
    \item $H$ has a tree decomposition whose largest bag has size at most 4 and such that some bag contains $x_1,\ldots,x_k$.
  \end{compactenum}
\end{lem}

\begin{proof}
  This proof very similar to the proof of Lemma~14 in \citet{dujmovic.joret.ea:planar}. Rather than duplicate every detail of that proof here, we focus on the differences and refer the reader to the original proof for the remaining details.

  The proof is by induction on the number of vertices of $N$.
  First note that $N'$ is a near-triangulation.  If $k=3$, set $R_i=P_i$ for each $i\in\{1,2,3\}$.  Otherwise, as in \citet{dujmovic.joret.ea:planar}, split $P_1,\ldots,P_k$ to partition $V(C)$ into three sets $R_1$, $R_2$, and $R_3$ such that each $F[R_i]$ is a path and each $R_i$ contains vertices from at most one of $P_1,\ldots,P_k$.

  Next, as in \citet{dujmovic.joret.ea:planar}, use Sperner's Lemma to find an inner face $\tau=v_1v_2v_3$ of $N'$ such that, $T$ contains disjoint vertical paths $Q_1,Q_2,Q_3$ such that each $Q_i$ begins at $v_i$, ends at some vertex in $R_i$, and whose internal vertices (if any) are contained in $V(N'-F)$.

  Let $\overline{Y}$ denote the subgraph of $N'$ consisting of vertices and edges of $Q_1$, $Q_2$, $Q_3$, and $\tau$.  Let $\overline{Y}^+$ denote the subgraph of $N$ consisting of the vertices and edges of $\overline{Y}$ plus the vertices and edges of every kite formed by a crossing between an edge of $G$ and an edge of $\overline{Y}$.

  We claim that, for each integer $i\ge 0$, $|V(Y^+)\cap V_i|\le 15$.  First observe that, since $Q_1,Q_2,Q_3$ are each vertical paths in $T$,  $\overline{Y}$ contains at most 3 vertices of $V_i$, each incident on at most two edges of $\overline{Y}$.  By \eqref{spar-distance}, any vertex $x\in V(\overline{Y}^+-\overline{Y})\cap V_i$, is incident on an edge $xy\in E(G)$ that crosses one of the at most six edges in $\overline{Y}$ having an endpoint in $V_i$.  These at most six edges have at most 12 endpoints.  Therefore $|V(\overline{Y}^+-\overline{Y})\cap V_i|\le 6\times 2=12$, so $|\overline{Y}^+\cap V_i|\le 12+3=15$.

  Finally, let $M$ and $M^+$ denote the subgraph of $G$ containing the edges and vertices of $Y$, respectively $Y^+$, and the edges and vertices of $F$.  The graph $M^+$ has some number of bounded faces, all contained in the interior of $F$. Some of the bounded faces of $M^+$ are kite faces. Let $F_1,\ldots,F_m$ be the non-kite bounded faces of $M^+$.

  We claim that, for each $i\in\{1,\ldots,m\}$, the boundary of $F_i$ is a cycle in $G'$.  Otherwise, some edge $vw$ contributes to the boundary of $F_i$ but is crossed by an edge $xy\in E(G)$. Then, $vw\not\in E(F)$ since no edge of $F$ is crossed by any edge of $G$. Therefore $vw\in E(\overline{Y}^+)$ so $xy\in E(Y^+)$. But then the only faces of $M^+$ incident to $vw$ are kite faces.  In particular $vw$ can not be incident the non-kite face $F_i$.

  Note that $M$ is a Mercedes graph and that each of the faces $F_1,\ldots,F_m$ is contained in a single wedge of $M$.   Let $Y^+ = \overline{Y}^+-F$. Therefore, $V(F_i)$ can be partitioned into at most three sets $P_1'$, $P_2'$, and $P_3'$ where $P_1'\subset V(Y^+)$, $P_2'\subseteq P_a$, $P_3'\subseteq P_b$ for some $a,b\in\{1,2,3\}$, and $F_i[P_j']$ is a path, for each $j\in\{1,2,3\}$.

  Finally, the subgraph $N_i$ of $G$ consisting of the edges and vertices of $G$ contained in $F_i$ or its interior does not contain one of the three vertices of $\tau$. Therefore, we can apply induction using the cycle $F_i$ and the partition $P_1',P_2',P_3'$ of $V(C_i)$ to obtain the desired $H$-partition and tree decomposition of $N_i$.

  The remainder of the proof finishes in the same way as the proof of Lemma~14 in \cite{dujmovic.joret.ea:planar}.  $P_1,\ldots,P_k$, and $V(Y^+-F)$ become elements of the $H$-partition.  Elements in each of the $H$-partitions of $N_1,\ldots,N_3$ that intersect $P_1,\ldots,P_k$, or $V(\overline{Y}^+-F)$ are discarded and all the resulting sets are combined to obtain an $H$-partition of $G$.  The desired tree decomposition of $G$ is obtained in exactly the same way as in the proof of Lemma~14 in \cite{dujmovic.joret.ea:planar}.

  The planarity of $H$ comes from two properties:
  \begin{enumerate}
    \item $G/\mathcal{P}$ and $G'\mathcal{P}$ are isomorphic.  This follows immediately from the fact that, for each edge $xy\in E(G-G')$, $x,y\in S_a$ for some $a\in V(H)$.  Indeed, $S_a=Y^+_a$ where $\overline{Y}_a$ contains the edge $vw\in E(G')$ that crosses $xy$.

    \item $G[\overline{Y}^+-F]$ is connected. To see why this is so, first observe that $\overline{Y}-F$ is connected, and then observe that every vertex of $\overline{Y^+}$ is either a vertex of $\overline{Y}$ or adjacent to a vertex of $\overline{Y}$.
  \end{enumerate}
  Since $G'$ is planar, this implies that $H=G'/\mathcal{P}$ is planar.\footnote{It is well-known and easy to see that any graph $\Delta'$ obtained by contracting an edge in a planar graph $\Delta$ is also planar.  Repeatedly applying this fact implies that $\Delta/\mathcal{P}$ is planar provided that $\Delta[P]$ is connected for every $P\in\mathcal{P}$.}
\end{proof}

Finally, we can complete the proof of \thmref{1-planar}.

\begin{proof}[Proof of \thmref{1-planar}]
  Using \lemref{induction}, the proof of \thmref{1-planar} is now straightforward. Given a 1-plane graph $G$, add edges to make it edge maximal so that it has an outer face $F=v_1v_2v_3$. Next, add a vertex $r$ adjacent to $v_1$, $v_2$, and $v_3$ to obtain an edge-maximal 1-plane graph $\overline{G}$ with one vertex $r$ of degree 3 on its outer face.

  Let $G'$ be the plane graph obtained by removing one spar from each kite of $\overline{G}$ and let $T$ be a BFS tree of $G'$ rooted at $r$.  Now apply \lemref{induction} with $G=\overline{G}$, $G'$, $F$, and $P_i=\{v_i\}$ for each $i\in\{1,2,3\}$.  This gives an $H$-partition $\{S_x:x\in V(H)\}$ of $\overline{G}-\{r\}\supseteq G$ in which $H$ is planar and has treewidth at most 3.

  Use the layering $\mathcal{L}=\{V_0',\ldots,V_h'\}$ where $V_i'=V_{2i}\cup V_{2i+1}$ for each $i\in\{0,\ldots,h\}$. That this is a layering of $G$ follows from \eqref{spar-distance}.  Since $|V_j\cap S_x|\le 15$ for every integer $i\ge 0$, $|V_i'\cap S_x|\le 30$ for every integer $i\ge 0$ and $x\in V(H)$.
\end{proof}

% \section{Consequences}
% \seclabel{consequences}
